Investigating soundscapes perception through acoustic scenes simulation

### 2018-09-04 lafaySimulatedScenePerception.tex
Relecture Mathias

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that the sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources on the notion of emotional quality of the scene, see~\cite{gozalo2015relationship,ricciardi2015sound}.

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download\footnote{Soundscapes, available online \url{https://archive.org/details/soundSimulatedUrbanScene}}.
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

To the best of our knowledge, only~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors shows that the addition or removal of the sources globally follows more social or semantic matters than their acoustical characteristics. A lack of diversity in terms of sound sources is nonetheless mentioned by the authors that limit the strength of the outcomes given in this study.

@@

The remaining of the paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. Experiments~1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section~\ref{sec:simulation}. Experiment~2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section~\ref{sec:modification}. Conclusions and discussion about future work follow in Section~\ref{sec:conclusion}.

@@

Once the typology and the set of sounds are available, an important design issue is the need of a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this may be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions. For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.

@@

Therefore, the selection interface considered in this study is text free and designed so as to force the user to rely on listening.

@@

Figure~\ref{fig:ssf} shows the interface used for the selection of events. Each circle corresponds to a sound class, the lowest level of abstraction (leaves) colored in grey. The spatial location of those circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the leaf classes, which are directly linked to a collection of samples.

@@

Each of those classes has a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sounds. The efficiency of this interface compared to several others designs have been evaluated and the outcomes are discussed by~\cite{lafay2016JAES}.

@@

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the two first experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):

@@

\begin{enumerate}
\item Simulation, where the user is asked to:
  \begin{itemize}
  \item  select sound classes,
  \item  give to each of them a name,
  \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
  \end{itemize}
\item Feedback : writing of a free form comment about the composed soundscape.
\end{enumerate}

@@

In addition, once the two sound scenes are completed, subjects are invited to:

@@

Before starting the first simulation, a 20-minute tutorial is given to the subjects in order to familiarize them with the simulation interface and the sound database. The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes). The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.  The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available online\footnote{Available online \url{http://soundthings.org/research/urbanSoundScape/XP2014}}.

@@

The assessment is done with a 7-point bipolar semantic scale going from -3 (non-ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, subjects must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to continue to the next scene.

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the experiments.}
\label{tab:acronyme}
\end{table}

@@

A set of features, upon which the analysis is conducted, is attached to each sound scene. A summary of those features (and the corresponding acronyms) is presented in Table~\ref{tab:acronyme}. In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).

@@

\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene, \ie~the sound classes that are present / absent from the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis. For instance, for the abstraction level 1, that includes 44 classes of sounds, the dimension is thus $44$ ($n = 44$).

\item \emph{structural features}: while \emph{SimScene} allows us to access a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels. To figure those out, we draw inspiration from the $L_{Aeq}$. In our case, it is a scalar computed from the signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
\end{itemize}

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the ideal and non-ideal scenes during the experiment 1.a. To do so, we consider if there exists significant differences between the ideal and non-ideal scenes for $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

@@

\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate if the sound levels ($L$, $L(E)$ and $L(T)$) impact the perceived pleasantness. To do so, we consider linear correlations between those features and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used in a given type of environment (ideal or non-ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used to simulate all the scenes, $c_k$ the number of classes used to simulate the scenes of a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used to simulate all the scenes, and $c_{jk}$ the number of time a class $j$ has been used to simulate the scenes of a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or} operator. Two scenes having similar source compositions will be close in such space. Using the Hamming distance allows us to take into account equally the presence and absence of classes. In order to measure the intrinsic ability of the space to discriminate between ideal and non-ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$). The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$. The $P@k$ is then the average of this ratio for all the items considered as search seeds.

@@

All statistical significance tests are conducted with a critical threshold of $\alpha=0.05$. For the V-test, considering that a large number of classes is tested, a Bonferroni correction is applied. For the $p$-value, if $p\geq0.05$, the value is reported. If $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.

@@

First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p<0.01$) than the non-ideal scenes.

@@

As expected, the sound level of the sources is indeed a pleasantness indicator, as the non-ideal scenes tend to be louder. This results is also an outcome of a large number of related works. We also notice that this difference of sound levels is significant for both events and textures.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non-ideal scenes. This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

However, sound level alone is not sufficient to fully differentiate between ideal and non-ideal scenes. In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$ remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($r=-0.32$, $p=0.06$) for ideal scenes.

@@

To sum up, for an unpleasant environment, sound levels --~especially those of events~-- negatively influence the perceived pleasantness. On the contrary, for a pleasant environment, none of the sound levels considered in the study seem to influence the perceived pleasantness.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (ideal or non-ideal), and 2) each involves distinct, independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness. We thus put forward the hypothesis that only the level of some of them can influence this perception. In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are made of.

@@

Nevertheless, we notice some differences with the results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes. The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors. Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.

@@

Regarding the sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are ideal scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the subjects who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted. Sound markers of non-ideal scenes are classes related to construction site (\emph{construction works}), or suggesting a strong traffic (\emph{horn}, \emph{siren}).

@@

Regarding the textures, 5 markers are identified. For the ideal scenes, those classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}). The marker classes for the non-ideal scenes are, as for the events, related to construction site (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).

@@

Concerning $ET$, the rank 5-precision is $76\%$ at abstraction level $0$ (the most abstract), and remains above $86\%$ for higher abstraction levels. Considering only the presence / absence of sound classes thus allows us to nicely discriminate between the two types of environments. We also notice that the less abstract (and therefore more precise) the description is, the more effective it is to predict agreement.

@@

The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector $S$ obtained when all the classes are considered ($ET$). $S$ being a boolean vector, the smaller the dimension, the lower the amount of information it can carry. Despite this, it appears that the rank 5-precision obtained with $E_m$ is equal – or superior – to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes. Reciprocally, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features. Thus, most of the semantic information allowing to differentiate between ideal and non-ideal scenes is included in the markers.

@@

On the contrary, the fact that we observe for the non-ideal scenes both significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$, shows that this is indeed the absolute level that matters for non ideal environments.

@@

In order to investigate deeper into the relation between the pleasantness of the ideal and non-ideal scenes and the sound markers, the audio waveform of the simulated scenes are regenerated without the classes identified as markers. To do so, ideal markers are removed from the ideal scenes, and non-ideal markers are removed from the non-ideal scenes. To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is conducted with a protocol close to the one considered in experiment 1.b.

@@

In order to create the scenes without markers with still some sound diversity and no absence of sound activity within a long period of the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}. Those classes are:

@@

In this section, we study the scores given by the subject while listening to the different types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers, see Figure~\ref{fig:xp2_Ascene}. The repeated measures ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the type environment (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@

These results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects, see Figure~\ref{fig:xp2A}. Our two hypotheses are thus verified :

@@

Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are simple to manipulate for subjects with little training, and possibly running on many types of hardware. While most processing frameworks that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulations can now be performed on modern web browsers using the webaudio liibrary.

@@

The experiments described here does not introduce yet another sound manipulation framework and is thus not designed to be extensible. Rather it is considered here to demonstrate the following:
\begin{itemize}
  \item the use of Webaudio javascript library is useful to design advanced experimental protocols that consider sound manipulation over the web
  \item the soundscape model considered is useful to question important research questions within the soundscape community  but simple enough to be grasped by novice participants.
\end{itemize}

@@

\begin{itemize}
\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture, provided that the sound databases are adapted to the cultural background of the cohort;
\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, the semantic nature of the sound classes used being known \emph{a priori} by the experimenter;
\item Simulated scenes can be analyzed without the need to annotate the sound scene to identify the sources, their occurrences in the simulated scene being directly available.
\end{itemize}

>>>>

We believe that the availability of a fine grain description of the sound stimuli is of great interest, because recent studies demonstrate that all sound sources do not contribute equally to the perception of the sound scene, see~\cite{defreville2004aactivity,lavandier2006contribution,guastavino2006ideal,nilsson2007soundscape,szeremeta2009analysis}. Thus, much attention is given to the specific contributions of the different sources to the notion of emotional quality of the scene, see~\cite{gozalo2015relationship,ricciardi2015sound}.

@@

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments (the resulting waveforms are available for download\footnote{Soundscapes, available online \url{https://archive.org/details/soundSimulatedUrbanScene}});
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

To the best of our knowledge, only~\cite{bruce2009development,bruce2014effects} considered the use of a simulator to question soundscape perception. They propose a tool that allows the user to modify a given soundscape by adding or removing specific sound sources, by changing the acoustic level of the sources as well as their spatial location. The authors show that the addition or removal of the sources globally follows social or semantic considerations more than their acoustical characteristics. A lack of diversity in terms of sound sources is nonetheless mentioned by the authors as a limiting factor to the strength of the outcomes given in this study.

@@

The remaining of this paper is organized as follows: the soundscape simulator \emph{simScene} is introduced in Section \ref{sec:simulator}. Experiments~1.a \emph{simulation} and  1.b \emph{pleasantness evaluation} are presented in Section~\ref{sec:simulation}. Experiment~2 \emph{pleasantness evaluation after modification of the scenes} is presented in Section~\ref{sec:modification}. Conclusions and discussion about future work follow in Section~\ref{sec:conclusion}.

@@

Once the typology and the set of sounds are available, an important design issue is the need for a suitable way to display the sound dataset to the user. Most browsing tools are based on keyword indexing; however, for sensory experiments that study the objectivation of a subject's mental representations, this may be problematic as the availability of a verbal description of the sound can influence the subject's choice, and potentially induce biases in the analysis conclusions. For example, a subject may automatically select sounds referenced as belonging to a \emph{park} environment to build a \emph{calm} soundscape, rather than focusing on their perception.

@@

Therefore, the selection interface considered in this study is text-free and designed so as to force the user to rely on listening.

@@

Figure~\ref{fig:ssf} shows the interface used for the selection of events. Each circle corresponds to a sound class, with the lowest level of abstraction (leaves) colored in grey. The spatial location of those circles is chosen according to the hierarchical organization of the sound database: sub-classes belonging to the same class are close to each others, and so on until the user reaches the leaf classes, which are directly linked to a collection of samples.

@@

Each of those classes has a representative sound chosen arbitrarily by the authors in order to provide the same sound each time the user clicks on the circle. The subject can browse the database by listening to those prototype sounds. The efficiency of this interface compared to several others designs has been evaluated and the outcomes are discussed by~\cite{lafay2016JAES}.

@@

Experiment 1 aims at using a simulation paradigm to investigate the specific influences of the various sound sources constituting urban soundscapes on the perceived pleasantness. For that, the first two experiments are planned as follows (\cf~Figure~\ref{fig:xp1_2}):

@@

\begin{enumerate}
\item Simulation, where the user is asked to:
  \begin{itemize}
  \item  select sound classes,
  \item  give each of them a name,
  \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
  \end{itemize}
\item Feedback : writing of a free form comment about the composed soundscape.
\end{enumerate}

@@

In addition, once the two sound scenes are completed, the subjects are invited to:

@@

Before starting the first simulation, a 20-minute tutorial is given in order to familiarize the subjects with the simulation interface and the sound database. The experiment is planned to last about two hours and a half, including breaks that the subjects are allowed to take.

@@

Among the 44 subjects, 40 succeeded, producing in the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes). The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.  The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available online\footnote{Available online \url{http://soundthings.org/research/urbanSoundScape/XP2014}}.

@@

The assessment is done with a 7-point bipolar semantic scale going from -3 (non-ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, the subjects must listen at least to the first 20 seconds of the stimuli. After the evaluation, they are free to continue to the next scene.

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Abbreviations              \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Abbreviations of features used in the analysis of the experiments.}
\label{tab:acronyme}
\end{table}

@@

A set of features, upon which the analysis is conducted, is attached to each sound scene. A summary of those features (and the corresponding abbreviations) is presented in Table~\ref{tab:acronyme}. In order to be consistent with the evaluation of experiment 1.b, features are not computed on the whole duration of the sequences but only on their 30-second reduced version used as stimuli for experiment 1.b (Section~\ref{sec:xp1b_plan}).

@@

\item \emph{semantic features}: we use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the classes of sounds involved in the scene, \ie~the sound classes that are present / absent from the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for the analysis. For instance, for the abstraction level 1, that includes 44 classes of sounds, the dimension is thus $44$ ($n = 44$).

\item \emph{structural features}: while \emph{SimScene} allows us to access a variety of information about the scene structure (such as the density of events), we only focus in this first study on the sound levels. To figure those out, we draw inspiration from the $L_{Aeq}$ measure. In our case, it is a scalar computed from the signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples, and only the set of texture samples.
\end{itemize}

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the ideal and non-ideal scenes during experiment~1.a. To do so, we consider if there exist significant differences between the ideal and non-ideal scenes for $\mathcal{A}_{scene}$ and $\mathcal{A}_{subject}$. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

@@

\item \emph{Influence of sound levels on perceived pleasantness}: the goal is to evaluate whether the sound levels ($L$, $L(E)$ and $L(T)$) affect the perceived pleasantness. To do so, we consider linear correlations between those features and $\mathcal{A}_{scene}$. The Pearson correlation coefficient is used for that purpose.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate whether specific classes are more frequently used in a given type of environment (ideal or non-ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used to simulate all the scenes, $c_k$ the number of classes used to simulate the scenes of a given type of environment $k$ (ideal or non-ideal), $c_j$ the number of times a class $j$ has been used to simulate all the scenes, and $c_{jk}$ the number of times a class $j$ has been used to simulate the scenes of a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or} operator. Two scenes having similar source compositions will be close in such a space. Using the Hamming distance allows us to take into account equally the presence and absence of classes. In order to measure the intrinsic ability of the space to discriminate between ideal and non-ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$). The $P@k$ computes the precision obtained after the $k$ closest items to a given seed item have been found. Formally, for each scene $s_i$ (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$. The $P@k$ is then the average of this ratio for all the items considered as search seeds.

@@

All statistical significance tests are conducted with a critical threshold of $\alpha=0.05$. For the V-test, considering that a large number of classes is tested, a Bonferroni correction is applied. For the $p$-value, if $p\geq0.05$, the value is reported; if $0.01\leq p<0.05$, we only report $p<0.05$, otherwise we report $p<0.01$.

@@

First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of an ideal scene. Four non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p<0.01$) than the non-ideal scenes.

@@

As expected, the sound level of the sources is indeed a pleasantness indicator, as the non-ideal scenes tend to be louder. This result is also an outcome of a large number of related works. We also notice that this difference of sound levels is significant for both events and textures.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ between ideal and non-ideal scenes. This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

However, sound level alone is not sufficient to fully differentiate between ideal and non-ideal scenes. In fact, $20\%$ of the ideal scenes have sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. However, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$ remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($r=-0.32$, $p=0.06$) for ideal scenes.

@@

To sum up, for an unpleasant environment, sound levels --\,especially those of events\,-- negatively influence the perceived pleasantness. On the contrary, for a pleasant environment, none of the sound levels considered in the study seem to influence the perceived pleasantness.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (ideal or non-ideal), and 2) each involves distinct, independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness. We thus put forward the hypothesis that only the level of some of them can influence this perception. In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are composed of.

@@

Nevertheless, we notice some differences with the results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes. The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors. Thus, in our representation of the world, sounds of public transportation would be positively connoted and tend to be more accepted than sounds of personal vehicles.

@@

Regarding sound events, 9 markers are identified for all abstraction levels. As shown on Figure~\ref{fig:soundsource}, classes related to human presence (\emph{male footsteps on concrete}, \emph{bicycle bell}), and of nature (\emph{animals, bird}, and \emph{bird song}) are ideal scenes markers as well as the \emph{church bell} class. This latter result may be due to the socio-cultural background of the subjects who are mostly European citizens. In fact, according to Schafer, a sound that is identified by a person as being an important element of his/her environment, is well accepted. Sound markers of non-ideal scenes are classes related to construction sites (\emph{construction works}), or suggesting intense traffic (\emph{horn}, \emph{siren}).

@@

Regarding textures, 5 markers are identified. For the ideal scenes, those are classes related to subdued or quiet ambiances (\emph{courtyard}, \emph{park}). The marker classes for the non-ideal scenes are, as for the events, related to construction sites (\emph{construction}, \emph{construction vehicle}), together with a class related to traffic (\emph{crossroads}).

@@

Concerning $ET$, the rank 5-precision is $76\%$ at abstraction level $0$ (the most abstract), and remains above $86\%$ for subsequent abstraction levels. Considering only the presence / absence of sound classes thus allows us to properly discriminate between the two types of environments. We also notice that the less abstract (and therefore more precise) the description is, the more effective it is to predict agreement.

@@

The dimension of the vectors $S$ for $E_m$ is lower than the dimension of vector $S$ for $E$, itself lower than the dimension of vector $S$ obtained when all the classes are considered ($ET$). $S$ being a boolean vector, the smaller the dimension, the lower the amount of information it can carry. Despite this, it appears that the rank 5-precision obtained with $E_m$ is equal –\,or superior\,– to the ones obtained with $E$ or $ET$, although only a partial information is used in that case to describe the scenes. Reciprocally, if the sound markers are not taken into account for the description ($E_{w/o,m}$), the performance is below the one achieved when considering only the textures as features. Thus, most if not all of the semantic information allowing to differentiate between ideal and non-ideal scenes is included in the markers.

@@

On the contrary, the fact that we observe for the non-ideal scenes both significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$, shows that it is indeed the absolute level that matters for non ideal environments.

@@

In order to investigate deeper into the relation between the pleasantness of ideal and non-ideal scenes and the sound markers, the audio waveforms of the simulated scenes are regenerated without the classes identified as markers. To do so, ideal markers are removed from the ideal scenes, and non-ideal markers are removed from the non-ideal scenes. To evaluate the impact on the perception of pleasantness caused by those modifications, a perceptual test is conducted with a protocol close to the one considered in experiment 1.b.

@@

In order to create the marker-less scenes with still some sound diversity and no absence of sound activity for long periods within the scene, only the sound classes of events of the first level of abstraction are removed, see Table~\ref{tab:markers}. Those classes are:

@@

In this section, we study the scores given by the subjects while listening to the different types of scenes, namely the ideal scenes with markers, non-ideal scenes with markers, ideal scenes without markers and non-ideal scenes without markers, see Figure~\ref{fig:xp2_Ascene}. The repeated measures ANOVA applied to $\mathcal{A}_{subject}$ shows a significant effect of the environment type (Type: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (marker: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

@@

Those results indicate that the removal of the markers indeed modify the perception of the scenes by the subjects, see Figure~\ref{fig:xp2A}. Our two hypotheses are thus verified :

@@

Considering sound scene synthesis or composition as part of an experimental protocol requires the availability of software resources that are simple to manipulate for subjects with little training, and possibly running on many types of hardware. While most processing frameworks that have been considered are developed in native language for efficiency purposes like Marsyas developed by~\cite{tzanetakis2000marsyas} or Clam developed by~\cite{amatriain2006clam}, many advanced sound manipulations can now be performed on modern web browsers using the webaudio library.

@@

The experiments described here does not introduce yet another sound manipulation framework and is thus not designed to be extensible. Rather it is considered here to demonstrate the following:
\begin{itemize}
  \item the use of Webaudio javascript library is useful to design advanced experimental protocols that consider sound manipulation over the web
  \item the considered soundscape model is useful to question important research questions within the soundscape community but simple enough to be grasped by novice participants.
\end{itemize}

@@

\begin{itemize}
\item The simulator can be deployed on a large scale via the Internet thanks to the web-based software architecture, provided that the sound databases are adapted to the cultural background of the test subject population;
\item Simulated scenes can be analyzed without the need to take into account the different mother tongues of the subjects, since the semantic natures of the used sound classes are known \emph{a priori} by the experimenter;
\item Simulated scenes can be analyzed without the need to annotate the sound scene to identify the sources, since their occurrences in the simulated scene are directly available.
\end{itemize}
